The Magnetorotational Instability Prefers Three Dimensions

### 2019-08-17 paper/paper.tex
first full pass through edits

% \documentclass[aps,prl,preprint,groupedaddress]{revtex4-1}
%\documentclass[aps,prl,preprint,superscriptaddress]{revtex4-1}
\documentclass[aps,prl,reprint,superscriptaddress]{revtex4-1}
\newcommand{\SSC}{S/S_{c}}
\newcommand{\mnras}{Monthly Notices of the Royal Astronomical Society}
\newcommand{\apjs}{The Astrophysical Journal Supplement Series}
\usepackage{graphicx}
% You should use BibTeX and apsrev.bst for references
% Choosing a journal automatically selects the correct APS
% BibTeX style file (bst file), so only uncomment the line
% below if necessary.
%\bibliographystyle{apsrev4-1}

@@

\begin{abstract}Ú
The magnetorotational instability (MRI) is a powerful instability where a weak magnetic can destabilize an electrically conducting fluid with a decreasing angular velocity shear.
The MRI is thought to be essential to the operation of astrophysical accretion disks, where the central object supplies a Keplerian angular velocity gradient.
The MRI may also be important in the internal shear layers of stars.
In stellar interiors, the shear can take a wide range to shear profiles, including near-critical, but with negligible dissipative the MRI may operate in a poorly studied regime where the shear is near critical but the dissipative forces are negligible.
Here, we consider the linear stability of three-dimensional perturbations to a nearly inviscid and ideal magnetofluid and show that the fastest growing modes are three dimensional when the instability is near onset.
These three dimensional modes are unstable at shear rates down to 0.10 $S_c$, the critical value in two dimensions, and they have growth rates exceeding the fastest two dimensional modes until the shear parameter is about $2.05 S_c$.
These modes are significant, as they suggest the possibility of direct dynamo action from the linear instability itself.
Furthermore, even when two dimensional modes have the fastest growth rates, significant numbers of three dimensional modes are unstable.
In such a system, the non-normal character of the MRI cannot be ignored anytime the shear is near critical.
Because evidence from weakly non-linear analysis suggests that the MRI saturates by pushing the shear rate to its critical value, non-normality is never unimportant for systems, like stars, that can rearrange their angular velocity profiles.
\end{abstract}

@@

The magnetorotational instability (MRI) is an extremely important instability in astrophysical fluid dynamics.
By changing the stabiliity criterion for differentially rotating flows from a negative angular \emph{momentum} gradient to a negative angular \emph{velocity} gradient, a weak magnetic field allows the Keplerian velocity profile due to a central point mass to drive turbulence \citep[e.g.][]{1998RvMP...70....1B}.
This discovery provided a pathway to explain the ubiquitous accretion onto compact objects at rates compatible with observations and may play a significant role in the formation of planets \citep[e.g.][]{2007Natur.448.1022J}.
In these objects, the gravitational field dominates the local plasma dynamics, and thus the MRI cannot significantly affect the background shear.
It must saturate by other means in these situations \citep{2018MNRAS.474.3451X}.
However, stars and liquid metal Taylor-Couette expereiments have differential rotation profiles driven by much weaker stresses.
Where the MRI is active in these flows, it can sature by pushing the background shear close to critical \citep{2017ApJ...841....1C,2017ApJ...841....2C}, analogous to the saturation of convection by relaxing to a nearly constant entropy profile.
Stellar interiors are at extremely high fluid and magnetic Reynolds numbers, but may potentially operate at or near the critical shear rate for the MRI\@.
This critical shear is set by the background magnetic field strength for axisymmetric perturbations in a channel of a given size.
Despite the extensive literature on the accretion disk case of strong shear and high Reynolds numbers and the liquid metal literature on weak shear and lower Reynolds numbers, even the linear dynamics of the MRI are not well studied in the weak shear, high Reynolds number regime.

Here, we investigate the stability of three dimensional perturbations near the two dimensional critical shear rate $S_c$ for a nearly invisicd, ideal MHD flow.
We find that the first modes to become destabilized are three-dimensional and thus act as a dynamo even in the absence of secondary instability.
These results strongly suggest that the non-normality of the linear operator driving the MRI are always important, even when axisymmetric modes dominate.
% \begin{itemize}
% \item 3D $\to$ dynamo
% \item stars, NSSL
% \item experiments
% \item Squire's theorem violation? Check MHD Squire, check S instead of Re...
% \item Do eigenvalue solution for B growth? i.e.\ check if it is a dynamo?
% \end{itemize}

We solve the linearized magnetohydrodynamic equations in rotating plane Couette geometry.
This corresponds to a Cartesian frame rotating with angular frequency $\Omega$ and a linear background shear, $V_{y} = Sx$. 
We cast the Navier-Stokes equation in a fairly standard form,
\begin{equation}
  \label{eq:mhd}
  \frac{D \mathbf{v}}{Dt} + f \hat{z} \times \mathbf{v} + S v_x \hat{z} + \mathbf{\nabla}{p} + \nu \mathbf{\nabla} \times \mathbf{\omega} = B_{0} \partial_{z} \mathbf{b},
\end{equation}
where
\begin{equation}
\omega = \mathbf{\nabla} \times \mathbf{v}, \quad \text{and} \quad \frac{D}{Dt} \ = \ \partial_{t} + S x \partial_{y}
\end{equation}
The main input parameters are the Coriolis parameter, $f = 2 \Omega$; the background shear rate, $S = dV_{y}/dx < 0$;  the vertical magnetic field $B_{0}$ (in Alfv\'{e}n units $\mu_{0} \rho_{0} = 1$); and the horizontal domain width $-d/2 \le x \le d/2$.
The accretion disk literature usually relates $S$ to $f$ via the $q = |S|/f$ parameter; for all calculations here, $q = 3/4$ (Keplerian).
We write the induction equation in terms of the $x$ component of the magnetic field,
\begin{equation}
  \label{eq:Bx}
  \frac{D b_x}{Dt} - B_0 \partial_z v_x + \eta (\partial_y j_z - \partial_z j_y) = 0
\end{equation}
and the $x$ component of the current density ($j_{x} = \partial_{y}b_{z} - \partial_{z} b_{y}$),
\begin{equation}
  \label{eq:Jx}
  \frac{D j_x}{Dt} - B_0 \partial_z \omega_x + S \partial_z B_x - \eta \nabla^2 j_x = 0
\end{equation}
We explicitly enforce the divergence constraint on the incompressible velocity field $v$ and the magnetic field,
\begin{equation}
  \label{eq:divu}
  \mathbf{\nabla} \cdot \mathbf{v} = \mathbf{\nabla} \cdot \mathbf{b} = 0.
\end{equation}
We consider the behavior of the MRI in a vertically ($z$) and horizontally ($y$) periodic channel of width $d = \pi$. The boundary conditions are impenetrable stress-free and perfectly conducting; $v_{x} = \omega_{y} = \omega_{z} = b_{x} = \partial_{x}j_{x} = 0$. The MRI is a weak-field instability; in the inviscid, ideal case the critical shear rate for instability in \textit{two dimensions} is
\begin{equation}
  \label{eq:Sc}
  S_c = -\frac{\pi^2 B_{0}^2}{f d^2}.
\end{equation}
Here, we use $\SSC$ as our control parameter.

We assume harmonic perturbations in $y$ and $z$, (e.g. pressure) $p = \hat{p}(x) e^{i(k_y y + k_z z) + \sigma t}$ . We use a complex-valued growth rate $\sigma = \gamma + i\omega$ with $\gamma$ and $\omega$ real. 
The system reduces to a $\sigma$-eigenvalue problem of ten first-order ODEs in $x$ with Dirichlet boundary conditions.
We pose and solve the system in the same formulation as equations~(\ref{eq:mhd}-\ref{eq:divu}) using the \emph{Dedalus} framework.

@@

Our two main results are: (I) When the first two-dimensional mode becomes unstable, there already exist three-dimensional modes with positive growth rate. (II) When the criticality becomes sufficiently large, the fastest-growing mode becomes purely two-dimensional. These results are pertinent in two ways. First, the MRI contains a ``Goldilocks regime" with possible direct dynamo action, and this regime likely most applies to stellar interiors. Second, our results accord with well-established results for accretion disk that expect primary two-dimensional linear modes.
Figure~\ref{fig:growth_rate}a shows the growth rates for four values of $\SSC$. 
At $\SSC = 1.02$, the maximum growth rate happens at $(k_y, k_z) \simeq (0.263, 0.447)$.
The first panel of figure~\ref{fig:growth_rate}a shows $\SSC = 0.640$, which contains only three-dimensional instability.
The overall critical shear for three-dimensional modes is $S_{c,3D} \simeq 0.
102 S_c$.

While surprising, that three-dimensional perturbations grow faster than two-dimensional ones can be predicted analytically.
We analyze equations~(\ref{eq:mhd}-\ref{eq:divu}) with the shear asymptotically close to the critical two-dimensional rate,  and compute the leading-order correction to the growth rate from three-dimensional effects as
\begin{widetext}
\begin{equation}
  \label{eq:asymp}
\sigma^{2} \ = -\frac{q\,B^{2}}{(q+1)} \left[ k_z^2 \left(\frac{d^2}{\pi^{2}} k_z^2-R^{2}\right) \ - \ \frac{\left(6+\pi ^2\right) q+\pi
   ^2-6}{12 } \, k_{y}^{2}  \right] \ + \  \ldots,
\end{equation}
\end{widetext}
where $S = S_c - \epsilon \sqrt{R}$, $k_{z} \sim \mathcal{O}(\epsilon)$, $\sigma \sim k_{y} \sim \mathcal{O}(\epsilon^{2})$.
The first term in equation~(\ref{eq:asymp}) is the result from a two-dimensional calculation.
The second term is negative definite and proportional to $k_{y}^2$, which leads to an ultraviolet catastrophe in the absence of higher-order effects.

Figure~\ref{fig:growth_rate}b compares the numerically growth rates for $\SSC=1.002$ between $0 \le k_y \le 0.2$ to this asymptotic approximation, showing that the numerical calculations accurately reproduce the analytic ones in the region where the latter is valid.
We provide the full spectrum for the most unstable parameter values for $\SSC = 1.02$ in figure~\ref{fig:growth_rate}c.
The single unstable mode is purely growing (orange dot), and its complex conjugate is also present (blue dot at $\omega = 0$), consistent with equation~(\ref{eq:asymp}).
The other stable modes (blue dots) are rotationally modified Alfven waves found in left- and right- going pairs, consistent with the analytic predictions for two-dimensional stability calculations.
%
%\onecolumngrid

\begin{figure*}[h!]
  \includegraphics[width=\textwidth]{fig_1.pdf}
  \caption{Growth rates for three-dimensional MRI modes. a) growth rates $\gamma$ over a grid of $k_y$ and $k_z$ for four values of supercriticality $\SSC$. The white contour highlights $\gamma = 0$; the red dots indicate the fastest growing mode. At $\SSC = 0.640$, there are no two dimensional modes. b) Growth rate vs $k_y$ for $k_z = ???$ at $\SSC = 1.002$ (highlighted by the blue line in a). The orange line gives the asymptotic result from equation~(\ref{eq:asymp}) and the blue line are the numerical results. The asymptotic form is valid as $\SSC \to 1$ and $k_y \ll 1$. c) The full discrete spectrum of the MRI for $(k_y, k_z) \simeq (0.263, 0.447)$. The unstable mode is shown in orange; all other modes are stable ($\gamma \le 0$).}
  \label{fig:growth_rate}
\end{figure*}
%\twocolumngrid
%
Numerical simulations have consistently shown axisymmetric modes dominating early evolution of the MRI before breaking down into MHD turbulence \citep{1995ApJ...440..742H,2018ApJ...853..174H,2019ApJS..241...26D}. 
These axisymmetric modes, called ``channel modes,'' are exact non-linear solutions of the governing equations with shearing-periodic boundary conditions in $x$, and thus must saturate via parasitic modes from physics beyond non-linear mode coupling \citep{1994ApJ...432..213G}.
Though we use impenetrable, stress-free boundary conditions that do not allow the unbounded growth of channel modes, the basic principle that axisymmetric disturbences grow fastest is not clearly caused by boundary conditions.
thus, we must reconcile our results with those earlier simulations.
In figure~\ref{fig:alpha}, we plot the angle $\alpha = \arctan k_y/k_z$ of the fastest growing mode with the $z$ axis as function of $\SSC$.
Above $\SSC \gtrsim 2.05$, $\alpha$ is zero, indicating that axisymmetric modes have the fastest growth rates.
For the fiducial run in \citet{1996ApJ...464..690H}, $\SSC \simeq 4.84$ (similar values are found in most works).
Thus, our theory predicts that the linear dynamics should be dominated by axisymmetric modes for the parameters studied in prior numerical simulations.
%
\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{alpha_vs_ssc_grid.pdf}
  \caption{Angle $\alpha$ of mode with respect to $z$ axis vs $\SSC$. The inset shows the transition from axisymmetric modes to three dimensional modes as $\SSC$ decreases below $2.05$. The shaded region indicates stability to all perturbations; the stability limit for three dimensional modes is $\SSC \simeq 0.102$.}
  \label{fig:alpha}
\end{figure}
%

However, even for $\SSC> 2.05$ there are significant swaths of unstable three-dimensional modes with growth rates comparable to the two-dimensional mode (figure~\ref{fig:growth_rate}a).
There are two significant implications for this.
First, the linear operator for the MRI, like other shearing systems, is non-normal; this effect appears whenever non-axisymmetric perturbations are allowed \citep[see][for a discussion relevant to the MRI]{1992MNRAS.255P..25K}.
Because of this, non-modal transient amplification occurs even for stable modes.
These effects can be strong enough to cause non-linear interactions and turbulence even when the entire flow is linearly stable.
While we are here concerned with systems linearly unstable to the MRI, the presence of three dimensional modes implies that linear, non-normal behavior could play a very strong role in the MRI near onset.

>>>>

% \documentclass[aps,prl,preprint,groupedaddress]{revtex4-1}
%\documentclass[aps,prl,preprint,superscriptaddress]{revtex4-1}
\documentclass[aps,prl,reprint,superscriptaddress]{revtex4-1}
\newcommand{\SSC}{S/S_{c}}
\newcommand{\mnras}{Monthly Notices of the Royal Astronomical Society}
\newcommand{\apjs}{The Astrophysical Journal Supplement Series}
\usepackage{graphicx}

% You should use BibTeX and apsrev.bst for references
% Choosing a journal automatically selects the correct APS
% BibTeX style file (bst file), so only uncomment the line
% below if necessary.
%\bibliographystyle{apsrev4-1}

@@

\begin{abstract}Ú
The magnetorotational instability (MRI) occurs when a weak magnetic field destabilizes a rotating electrically conducting fluid with an inwardly increasing angular velocity shear.
The MRI is considered essential to the operation of astrophysical accretion disks, with the central object supplying a Keplerian rotation profile.
The MRI should also exist for the internal shear layers of stars.
Stellar-interior shears can take a wide range of profiles, including near-critical, but with negligible dissipation. 
Here, we show that the fastest growing unstable modes of an ideal magnetofluid are three dimensional provided the shear rate, $S$, is close to the two-dimensional onset value, $S_c$.
Three-dimensional modes are unstable down to $S \approx 0.10 S_c$, and dominate the two-dimensional modes until $S \approx 2.05 S_c$.
Significant numbers of fast three dimensional modes remain well past $2.05 S_{c}$. 
These finding are significant in three ways. 
First, evidence from weakly nonlinear theory suggests that the MRI saturates by pushing the shear rate to its critical value. 
This can happen for systems, like stars, that can rearrange their angular velocity profiles.
Second, the non-normal character and large transient growth of the MRI should be important anytime strong three-dimensionality exists.
Finally, three-dimensional growth suggest the possibility of direct dynamo action driven from the linear instability itself.
\end{abstract}

@@

The magnetorotational instability (MRI) is an extremely important in astrophysical fluid dynamics.
By changing the stabiliity criterion for differentially rotating flows from a negative angular \emph{momentum} gradient to a negative angular \emph{velocity} gradient, a weak magnetic field allows the Keplerian velocity profile due to a central point mass to drive turbulence \citep[e.g.][]{1998RvMP...70....1B}.
This discovery explained the ubiquitous accretion onto compact objects at rates compatible with observations, and may also influence the formation of planets \citep[e.g.][]{2007Natur.448.1022J}.
In disks, the gravitational field dominates the local plasma dynamics, and thus the MRI cannot significantly affect the background shear; it must saturate by other means \citep{2018MNRAS.474.3451X}.
However, stars and liquid metal Taylor-Couette experiments have differential rotation profiles driven by much weaker stresses.
Where the MRI is active in these flows, it can saturate by pushing the background shear close to critical \citep{2017ApJ...841....1C,2017ApJ...841....2C}, analogous to convection mixing entropy.
Stellar interiors have extremely high fluid and magnetic Reynolds numbers, but may potentially operate at or near the critical shear rate for the MRI.
This critical shear is set by the background magnetic field strength for axisymmetric perturbations in a channel of a given size.
Despite the extensive literature on the accretion disk case of strong shear and high Reynolds numbers and the liquid metal literature on weak shear and lower Reynolds numbers, even the linear dynamics of the MRI are not well studied in the weak shear, high Reynolds number regime.

Here, we investigate the stability of three dimensional perturbations near the two dimensional critical shear rate $S_c$ for a nearly invisicd, ideal MHD flow.
We find that the first destabilized modes are three-dimensional, and thus can act as a dynamo even in the absence of secondary instability.
These results strongly suggest that the non-normality of the linear operator driving the MRI are always important, even when axisymmetric modes dominate.
% \begin{itemize}
% \item 3D $\to$ dynamo
% \item stars, NSSL
% \item experiments
% \item Squire's theorem violation? Check MHD Squire, check S instead of Re...
% \item Do eigenvalue solution for B growth? i.e.\ check if it is a dynamo?
% \end{itemize}

We solve the linearized magnetohydrodynamic equations in rotating plane-Couette geometry.
This corresponds to a Cartesian frame rotating with angular frequency $\Omega$ and a linear background shear, $V_{y} = Sx$. 
We cast the Navier-Stokes equation in a fairly standard form,
\begin{equation}
  \label{eq:mhd}
  \frac{D \mathbf{v}}{Dt} + f \hat{z} \times \mathbf{v} + S v_x \hat{z} + \mathbf{\nabla}{p} + \nu \mathbf{\nabla} \times \mathbf{\omega} = B_{0} \partial_{z} \mathbf{b},
\end{equation}
where
\begin{equation}
\omega = \mathbf{\nabla} \times \mathbf{v}, \quad \text{and} \quad \frac{D}{Dt} \ = \ \partial_{t} + S x \partial_{y}
\end{equation}
The main input parameters are the Coriolis parameter, $f = 2 \Omega$; the background shear rate, $S = dV_{y}/dx < 0$;  the vertical magnetic field $B_{0}$ (in Alfv\'{e}n units $\mu_{0} \rho_{0} = 1$); and the horizontal domain width $-d/2 \le x \le d/2$.
Accretion disk modelling usually considers the Rossby number $q = -S/f < 1$; $q=1$ corresponds to purely hydrodynamical Rayleigh unstable shear.
For all calculations here, we use $q = 3/4$ (Keplerian).
We write the induction equation in terms of the $x$ component of the magnetic field,
\begin{equation}
  \label{eq:Bx}
  \frac{D b_x}{Dt} - B_0 \partial_z v_x + \eta (\partial_y j_z - \partial_z j_y) = 0
\end{equation}
and the $x$ component of the current density ($j_{x} = \partial_{y}b_{z} - \partial_{z} b_{y}$),
\begin{equation}
  \label{eq:Jx}
  \frac{D j_x}{Dt} - B_0 \partial_z \omega_x + S \partial_z B_x - \eta \nabla^2 j_x = 0
\end{equation}
We explicitly enforce the divergence constraint on the incompressible velocity field $v$ and the magnetic field,
\begin{equation}
  \label{eq:divu}
  \mathbf{\nabla} \cdot \mathbf{v} = \mathbf{\nabla} \cdot \mathbf{b} = 0.
\end{equation}
We consider the behavior of the MRI in a vertically ($z$) and horizontally ($y$) periodic channel of width $d = \pi$. The boundary conditions are impenetrable stress-free and perfectly conducting; $v_{x} = \omega_{y} = \omega_{z} = b_{x} = \partial_{x}j_{x} = 0$. The MRI is a weak-field instability; in the inviscid, ideal case the critical shear rate for instability in \textit{two dimensions} is
\begin{equation}
  \label{eq:Sc}
  S_c = -\frac{\pi^2 B_{0}^2}{f d^2}.
\end{equation}
We use $\SSC$ as our instability control parameter.

We assume harmonic perturbations in $y$ and $z$, (e.g. pressure) $p = \hat{p}(x) e^{i(k_y y + k_z z) + \sigma t}$ . 
We use a complex-valued growth rate $\sigma = \gamma + i\omega$ with $\gamma, \omega$ both real. 
The system reduces to a $\sigma$-eigenvalue problem of 10 first-order ODEs in $x$ with Dirichlet boundary conditions.
We pose and solve the system in the same formulation as equations~(\ref{eq:mhd}-\ref{eq:divu}) using the \emph{Dedalus} framework.

@@

Our two main results are: (I) When the first two-dimensional mode becomes unstable, there already exist three-dimensional modes with positive growth rate. (II) When the criticality becomes sufficiently large, the fastest-growing mode becomes purely two-dimensional. These results are pertinent in two ways. First, the MRI contains a substantial ``Goldilocks regime" with possible direct dynamo action, and this regime likely most applies to stellar interiors. Second, our results accord with well-established results for accretion disks that expect primary two-dimensional linear modes.
Figure~\ref{fig:growth_rate}a shows the growth rates for four values of $\SSC$. 
At $\SSC = 1.02$, the maximum growth rate ($\sigma = ??$) happens at $(k_y, k_z) \simeq (0.263, 0.447)$.
The first panel of figure~\ref{fig:growth_rate}a shows $\SSC = 0.640$, which contains only three-dimensional instability.
The overall critical shear for three-dimensional modes is $S_{c,3D} \simeq 0.102 S_c$.

While surprising, preferred three-dimensional modes can be predicted analytically.
We analyze equations~(\ref{eq:mhd}-\ref{eq:divu}) with the shear asymptotically close to the critical two-dimensional rate, and compute the leading-order correction to the growth rate from three-dimensional effects assuming where $S = S_c - \epsilon \sqrt{R}$, $k_{z} \sim \mathcal{O}(\epsilon)$, $\sigma \sim k_{y} \sim \mathcal{O}(\epsilon^{2})$, 
%\begin{widetext}
\begin{equation}
  \label{eq:asymp}
\sigma^{2} =  -\frac{q\,B_{0}^{2}}{(q+1)} \left[ k_z^2 \left( \frac{d^2}{\pi^{2}} k_z^2 - R^{2}\right)  -  \Upsilon \, k_{y}^{2}  \right] +  \ldots,
\end{equation}
where
\begin{equation}
\Upsilon \ = \ \frac{\left(6+\pi ^2\right) q+\pi^2-6}{12 }\  \approx \ 1.31 \ \  \text{for}  \ \  q = \frac{3}{4}.
\end{equation}
%\end{widetext}
The first term in equation~(\ref{eq:asymp}) results from the two-dimensional calculation.
The second term is negative definite and leads ultraviolet divergence in the absence of higher-order effects.

Figure~\ref{fig:growth_rate}b compares the numerical growth rates for $\SSC=1.002$ between $0 \le k_y \le 0.2$ to the asymptotic approximation, showing good agreement where the latter is valid.
Figure~\ref{fig:growth_rate}c shows the full spectrum for $\SSC = 1.02$.
The plot shows a purely growing/decaying complex-conjugate pair (orange/blue dots on the real axis) consistent with equation~(\ref{eq:asymp}).
The other stable modes (blue dots) are rotationally modified Alfv\'{e}n waves found in left- and right-going pairs, consistent with the analytic predictions for two-dimensional stability calculations.
%\onecolumngrid
%
\begin{figure*}[h!]
  \includegraphics[width=\textwidth]{fig_1.pdf}
  \caption{Growth rates for three-dimensional MRI modes. a) growth rates $\gamma$ over a grid of $k_y$ and $k_z$ for four values of supercriticality $\SSC$. The white contour highlights $\gamma = 0$; the red dots indicate the fastest growing mode. At $\SSC = 0.640$, there are no two dimensional modes. b) Growth rate vs $k_y$ for $k_z = ???$ at $\SSC = 1.002$ (highlighted by the blue line in a). The orange line gives the asymptotic result from equation~(\ref{eq:asymp}) and the blue line are the numerical results. The asymptotic form is valid as $\SSC \to 1$ and $k_y \ll 1$. c) The full discrete spectrum of the MRI for $(k_y, k_z) \simeq (0.263, 0.447)$. The unstable mode is shown in orange; all other modes are stable ($\gamma \le 0$).}
  \label{fig:growth_rate}
\end{figure*}
%\twocolumngrid
%
Numerical simulations consistently show axisymmetric modes dominating the early evolution of the MRI before breaking down into 3D MHD turbulence \citep{1995ApJ...440..742H,2018ApJ...853..174H,2019ApJS..241...26D}. 
These ``channel modes'' are exact non-linear solutions for $x$-shearing-periodic domains, and can only saturate via parasitic shear instabilities \citep{1994ApJ...432..213G}.
We use impenetrable, stress-free boundary conditions that are more applicable to stellar interiors. 
But even in disks, any kind of finite radial extent will cutoff the unbounded growth of channel modes. 
We reconcile our results with earlier simulations by showing that the MRI indeed prefers two dimensions for the larger values of  criticality found in disk simulations. 
Figure~\ref{fig:alpha} shows the phase angle $\phi = \arctan k_y/k_z$ of the fastest growing mode as function of $\SSC$.
Above $\SSC \gtrsim 2.05$ $\phi=0$, indicating that axisymmetric modes have the fastest growth rates.
For the fiducial run in \citet{1996ApJ...464..690H}, $\SSC \simeq 4.84$ (most works use similar values).
Thus, we predict axisymmetric modes should dominate the linear dynamics for parameters studied in prior numerical simulations.
%
\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{alpha_vs_ssc_grid.pdf}
  \caption{Angle $\alpha$ of mode with respect to $z$ axis vs $\SSC$. The inset shows the transition from axisymmetric modes to three dimensional modes as $\SSC$ decreases below $2.05$. The shaded region indicates stability to all perturbations; the stability limit for three dimensional modes is $\SSC \simeq 0.102$.}
  \label{fig:alpha}
\end{figure}
%

Even for $\SSC> 2.05$ there are significant swaths of unstable three-dimensional modes with growth rates comparable to the maximum (figure~\ref{fig:growth_rate}a).
Also, non-normality generically accompanies non-axisymmetry in shearing systems \citep[see][for a discussion relevant to the MRI]{1992MNRAS.255P..25K}.
In non-normal systems, transient amplification can occur for stable modes, and can even cause turbulence.
Therefore significant three dimensionality likely implies important non-normal behavior near onset.
